\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-31 paper/ms.tex
updating to latest version
Flagged variables that are reused (which can lead to confusion of reader).  Discussed real terms.

\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
\altaffiltext{\guggenheim}{Guggenheim Fellow}

@@

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The data are simulated from a quasiperiodic GP with the kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\citep{Wilson:2013}
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
exactly reproduce the structure of this process.
That being said, we demonstrate that robust inferences can be made about
$P_\mathrm{true}$ even with this effective model.

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

\newpage
\input{figures/transit/transit-params}

@@

>>>>

\edef \guggenheim {\arabic{affilcounter}}\stepcounter{affilcounter}
\altaffiltext{\guggenheim}{Guggenheim Fellow}

@@

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$. \response{X was used to represent independent
parameter matrix in definition of likelihood.}
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
then these terms only add one to the rank of $U$ and $V$.  We use this fact
to simplify the algebra when real components are included:  if $J_{real}$
and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then 
the $U$, $V$, and $X$ matrices have size $N \times (J_{real}+2J_{complex})$.

@@

The data are simulated from a quasiperiodic GP with the kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\citep{Wilson:2013}
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially. \response{$\alpha$
means the parameter vector in definition of likelihood.}
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
exactly reproduce the structure of this process.
That being said, we demonstrate that robust inferences can be made about
$P_\mathrm{true}$ even with this effective model.

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$. \response{a,b,c already have different meaning
in celerite kernel.}
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters. \response{A was used for diagonal of
covariance matrix.}
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

\newpage

\response{Table 5 has two parameter with name "b":  the celerite coefficient,
and the impact parameter.}

\input{figures/transit/transit-params}

@@

\response{The D in Table 6 may be confused with the D used in the Cholesky solver.}
